Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-03-02 paper/paper1/ms.tex
Minor tweaks to the introduction

The metaphorical holy grail of astronomical spectroscopy is a function that takes in an observed astronomical spectrum and reports back the postition, amplitude, width, and shape of all of its spectral lines, automatically, accurately, and precisely.  The function would go further, to report back the systemic radial velocity (RV) and rotational broadening ($v\sin{i}$).  Finally the function would---in an \emph{Inception}-like manner---provide what it believes to be the interpretable generating function that produced the data, so that we may gain insights on future examples of this or other stars.  Solving this problem is hard, for at least four reasons.  First, the spectral lines may overlap, and so the assignment of one line may be partially degenerate with the assignment of some other adjacent line.  Second, extremely wide line wings blend into the continuum, such that the placement of the continuum level may become ill-defined.  Third, the extent of line blending and realized line shape depends strongly on the spectral resolution of the spectrograph, the rotational broadening of the star, and possibly the instrumental configuration at the time of observation.  Finally, telluric absorption lines commingle with the astronomical spectral lines of interest, censoring some spectral regions entirely, or partially confounding other lines with chance allignments.

@@

But many new and important questions in the fields of stars and exoplanets aspire to reach the margins of what the entire dataset can inform.  In particular, data from high-\emph{grasp} \'echelle spectrographs possess simultaneously high spectral resolving power and high bandwidth, yielding tens of thousands or possibly millions of independent spectral resolution elements for each star, substar, or exoplanet.  Those applications that seek to gain signal by ``stacking'' spectral lines or cross-correlating templates can hypothetically gain huge boosts in the accessible signal-to-noise ratio compared to a single or few lines.  Most manual and semi-automated methods cannot take advantage of the entire spectral bandwidth, or rely on exact knowledge of the underlying templates and may fail to achieve the hypothetical promise of these high-bandwidth spectrographs \citep{2020AJ....160..198H}.

For example, exoplanet cross-correlation spectroscopy \citep{2013MNRAS.436L..35B} hinges on accurate molecular spectral templates to detect and characterize the atmospheres of exoplanets.  Imperfections in these templates can mute the perceived signal strength of these atmospheric features \citep{2015A&A...575A..20H}.  
In extreme precision radial velocity (EPRV) applications, cross-correlation methods work, but addressing telluric absorption at the $\mathrm{cm/s}$ level may require joint modeling of the star and the Earth's atmospheric absorption before convolution with an instrumental kernel \authorcomment1{speakerdeck.com/dwhgg/telluric-line-hack-week-wrap-up}).

In the case of Doppler imaging, an accurate underlying template is needed to detect longitudinally symmetric structures such as polar spots \citep{1983PASP...95..565V, roettenbacher16} or zonal bands \citep{Crossfield14,2021ApJ...906...64A}.

Existing open-source frameworks have overcome some of these challenges, or have been purpose-built for specialized applications. 
These frameworks include \texttt{ROBOSPECT} \citep{2013PASP..125.1164W}, \texttt{specmatch} \citep{2015PhDT........82P}, \texttt{specmatch-emp} \citep{2017ApJ...836...77Y}, \texttt{wobble} \citep{2019AJ....158..164B}, \texttt{starfish} \citep{czekala15}, \texttt{sick} \citep{2016ApJS..223....8C}, \texttt{psoap} \citep{2017ApJ...840...49C}, \texttt{FAL} \authorcomment1{Cargile et al. in prep}, CHIMERA \citep{2015ApJ...807..183L}, the \texttt{Cannon} \citep{2017ApJ...836....5H},  \texttt{MINESweeper} \citep{2020ApJ...900...28C}, and recently \texttt{ExoJAX} \citep{2022ApJS..258...31K}.  
The design of these frameworks necessarily have to make a choice in the bias-variance tradeoff: is the tool more \emph{data}-driven or more \emph{model}-driven?  The statistical tradeoff can be viewed as a concession in physical self-consistency for model flexibility: more or fewer parameters.

A key new enabling technology breaks the historic limits of data-model comparisons for astronomical spectroscopy.  Automatic differentiation \citep[``autodiff'',][]{2015arXiv150205767G} and the backpropagation algorithm has revolutionized machine learning and neural network architecture design, and is increasingly applied in astrophysical data analysis contexts, \emph{e.g.} kernel phase coronography with \texttt{poppy} \citep{2021ApJ...907...40P}, and exoplanet orbit fitting with \texttt{exoplanet} \citep{2021JOSS....6.3285F}.  Of the spectroscopy frameworks mentioned above, \texttt{wobble} \citep{2019AJ....158..164B} and \texttt{ExoJAX} \citep{2022ApJS..258...31K} employ autodiff technology.  \texttt{wobble} treats each pixel as a tunable control point, producing $\sim10^5$ parameters for a modern stellar spectrum.  The \texttt{ExoJAX} framework has only $\sim$dozens of tunable parameters that describe the fundamental physical properties controlling a brown dwarf atmosphere.  These two autodiff-aware frameworks span the extreme ends of non-parametric and parametric modeling for spectroscopy.

In this paper we show that autodiff-aware semi-empirical models offer a cozy middle ground: informed from self-consistent models, but refined with data.  In Section \ref{methodology} we show how to supplant the computationally expensive radiative transfer, by cloning pre-computed synthetic spectral models.  In Section \censor{XX} we describe how to fine-adjust the cloned models to data in a transfer-learning step.  Section \censor{XX} exhibits some injection-recovery type tests on simulated data of known properties.  In Section \censor{XX} we show real-world data from high-grasp near-IR spectrographs.  Finally we discuss many conceivable extensions to the framework, and directions for future work in Section \censor{xx}.

>>>>

The metaphorical holy grail of astronomical spectroscopy is a function that takes in an observed astronomical spectrum and reports back the postition, amplitude, width, and shape of all of its spectral lines, automatically, accurately, and precisely.  The function would go further. It would report back the systemic radial velocity (RV) and rotational broadening ($v\sin{i}$).  Finally the function would---in an \emph{Inception}-like manner---provide what it believes to be the interpretable generating function that produced the data, so that we may gain insights on future examples of this or other stars.  Solving this problem is hard, for at least four reasons.  First, the spectral lines may overlap, and so the assignment of one line may be partially degenerate with the assignment of some other adjacent line.  Second, extremely wide line wings blend into the continuum, such that the placement of the continuum level may become ill-defined.  Third, the extent of line blending and realized line shape depends strongly on the spectral resolution of the spectrograph, the rotational broadening of the star, and possibly the instrumental configuration at the time of observation.  Finally, telluric absorption lines commingle with the astronomical spectral lines of interest, censoring some spectral regions entirely, or partially confounding other lines with chance allignments.

@@

But many new and important questions in the fields of stars and exoplanets aspire to reach the margins of what the entire dataset can inform.  In particular, data from high-\emph{grasp} \'echelle spectrographs possess simultaneously high spectral resolving power and high bandwidth, yielding tens of thousands or possibly millions of independent spectral resolution elements for each star, substar, or exoplanet.  Those applications that seek to gain signal by ``stacking'' spectral lines or cross-correlating with templates can hypothetically gain huge boosts in the accessible signal-to-noise ratio compared to a single or few lines.  Most manual and semi-automated methods cannot take advantage of the entire spectral bandwidth, or rely on exact knowledge of the underlying templates and may fail to achieve the hypothetical promise of these high-bandwidth spectrographs \citep{2020AJ....160..198H}.

For example, exoplanet cross-correlation spectroscopy \citep{2013MNRAS.436L..35B} hinges on accurate molecular spectral templates to detect and characterize the atmospheres of exoplanets.  Imperfections in these templates can mute the perceived signal strength of these atmospheric features \citep{2015A&A...575A..20H}.

In extreme precision radial velocity (EPRV) applications, cross-correlation methods work \citep{2018A_A...620A..47D}, but have many limitations \citep{2022arXiv220110639Z}.  Addressing telluric absorption at the $\mathrm{cm/s}$ level may require joint modeling of the star and the Earth's atmospheric absorption before convolution with an instrumental kernel\authorcomment1{speakerdeck.com/dwhgg/telluric-line-hack-week-wrap-up}.  This telluric joint modeling capability does not yet exist at a precision that can meet these strenuous demands.

In the case of Doppler imaging, an accurate underlying template is needed to detect longitudinally symmetric structures \citep{1983PASP...95..565V,2021arXiv211006271L} such as polar spots \citep{roettenbacher16} or zonal bands \citep{Crossfield14,2021ApJ...906...64A}.

Existing open-source frameworks have overcome some of these challenges, or have been purpose-built for specialized applications. 
These frameworks include \texttt{ROBOSPECT} \citep{2013PASP..125.1164W}, \texttt{specmatch} \citep{2015PhDT........82P}, \texttt{specmatch-emp} \citep{2017ApJ...836...77Y}, \texttt{wobble} \citep{2019AJ....158..164B}, \texttt{starfish} \citep{czekala15}, \texttt{sick} \citep{2016ApJS..223....8C}, \texttt{psoap} \citep{2017ApJ...840...49C}, \texttt{FAL} \authorcomment1{Cargile et al. in prep}, CHIMERA \citep{2015ApJ...807..183L}, the \texttt{Cannon} \citep{2017ApJ...836....5H},  \texttt{MINESweeper} \citep{2020ApJ...900...28C}, and recently \texttt{ExoJAX} \citep{2022ApJS..258...31K}.  
The design of these frameworks necessarily have to make a choice in the bias-variance tradeoff: is the tool more \emph{data}-driven or more \emph{model}-driven?  The statistical tradeoff can be viewed as a concession in physical self-consistency for model flexibility: more or fewer parameters; more \emph{accurate} or more \emph{precise}.

A key new enabling technology breaks these classical tradeoffs in data-model comparisons for astronomical spectroscopy.  Automatic differentiation \citep[``autodiff'',][]{2015arXiv150205767G} and its affiliated backpropagation algorithm has revolutionized machine learning and neural network architecture design, and is increasingly applied in astrophysical data analysis contexts, \emph{e.g.} kernel phase coronography with \texttt{poppy} \citep{2021ApJ...907...40P}, and exoplanet orbit fitting with \texttt{exoplanet} \citep{2021JOSS....6.3285F}.  Of the spectroscopy frameworks mentioned above, the \texttt{TensorFlow}-based \citep{tensorflow2015-whitepaper} \texttt{wobble} and the \texttt{JAX}-based \citep{jax2018github} \texttt{ExoJAX} employ autodiff technology.  \texttt{wobble} treats each pixel as a tunable control point, producing $\sim10^5$ parameters for a modern stellar spectrum.  The \texttt{ExoJAX} framework has only $\sim$dozens of tunable parameters that describe the fundamental physical properties controlling a brown dwarf atmosphere.  These two autodiff-aware frameworks span the extreme ends of non-parametric and parametric modeling for spectroscopy.

In this paper we show that autodiff-aware semi-empirical models offer an appealing middle ground: informed from self-consistent models, but refined with data.  In Section \ref{methodology} we introduce the \emph{blas\'e} forward-model design and its \texttt{PyTorch}-based \cite{2019arXiv191201703P} implementation.  We show how to supplant the computationally expensive radiative transfer by cloning pre-computed synthetic spectral models.  In Section \censorbox{XX} we describe how to fine-adjust the cloned models to data in a transfer-learning step.  Section \censorbox{XX} exhibits some injection-recovery type tests on simulated data of known properties.  In Section \censorbox{XX} we show real-world data from high-grasp near-IR spectrographs.  Finally we discuss many conceivable extensions to the framework, and directions for future work in Section \censorbox{XX}.
